---
title: "Background"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(janitor)
library(magrittr)
library(reactable)
library(tidycensus)
library(tidyverse)
```

On November 13, 2022, three UVA football players were shot and killed on campus. Devin Chandler, Lavel Davis Jr. and D’Sean Perry lost their lives to an act of gun violence. The rate of violent crime in Charlottesville increased by 30% from 2021 to 2022, and in the early months of 2023, Cville police had already investigated five homicide cases; compared to zero cases in 2021 ([Cville Community Safety Report](https://prescouncil.president.virginia.edu/sites/g/files/jsddwu616/files/2024-01/Final%20Community%20Safety%20Working%20Group%20Report_24.pdf)). This document attempts to recognizing the commonality of our experiences but also includes an honest recognition of our differences. How can UVA and the wider Charlottesville and Albemarle county community work together to end gun violence in our area?

# Who are the stakeholders?

Before we even start looking at the data, we need to consider the various stakeholders in this project. What are they looking to understand from the data? What are they looking to do with insights from the data? Or with the actual data, itself. What biases might folks have? What questions are they trying to answer? Someone may be looking to learn how gun violence has changed over the years, while someone else may be interested in understanding how to prevent gun violence, while others are looking for the structural causes. These are all things to keep in mind.

Ultimately the residents of Charlottesville and Albemarle are the largest stakeholders. These folks ARE the data. They have the lived experiences, and are impacted in various ways. Their voices should be centered.

# What is gun violence?

The CDC defines a <b>firearm injury</b> as:

> "a gunshot wound or penetrating injury from a weapon that uses a powder charge to fire a projectile. Weapons that use a powder charge include handguns, rifles, and shotguns. Injuries from air- and gas-powered guns, BB guns, and pellet guns are not considered firearm injuries as these types of guns do not use a powder charge to fire a projectile." ([CDC](https://www.cdc.gov/violenceprevention/firearms/fastfact.html))

Types of firearm injuries include:

-   Intentionally self-inflicted
-   Unintentional (accidental injuries that happen while someone is cleaning or playing with a firearm)
-   Interpersonal violence (homicide or assault)
-   Legal intervention (violence inflicted by law enforcement on duty)
-   Undetermined intent (not enough information to determine)

---

# What are the national trends? 

---

# What is the prevalence in our region? 

The chart below includes the number and rate of firearm-related deaths among Blue Ridge Health District residents by Intent over the past 5 combined years. The Blue Ridge Health District includes Charlottesville City, Albemarle, Greene, Louisa, Fluvanna, and Nelson County. The leading cause of death by gun violence in the Blue Ridge Health District is suicide. 

```{r vdh_intent, echo=FALSE, message=FALSE}

fai_intent <- read_csv("data/raw/vdh-pud-firearm-deaths-by-district-intent.csv") 

fai_intent %>%
  clean_names() %>%
  filter(str_detect(patient_health_district, "Blue Ridge")) %>%
  arrange(desc(firearm_deaths)) %>%
  reactable(
    defaultColDef = colDef(
    header = function(value) gsub("_", " ", value, fixed = TRUE),
    minWidth = 70,
    headerStyle = list(background = "#f7f7f8")
  )) %>%
  reactablefmtr::add_source("Source: https://data.virginia.gov/dataset/vdh-pud-firearm-deaths-by-district-intent", align = "right", font_style = "italic", font_size = 10, font_color = "grey")
```

---


# Where are people getting guns?

## Licensed Dealers

```{r dealers_tidycensus, results='hide'}
alb_sf <- get_acs(
  geography = "county",
  state = "51",
  variables = "B01001_001",
  survey = "acs5",
  geometry = TRUE,
  year = 2022) %>%
  filter(NAME == "Albemarle County, Virginia")
```

The ATF maintains a list of all licensed firearms dealers. The below map shows authorized dealers in our region.

```{r dealers, echo = FALSE, message = FALSE, warning = FALSE}

atf_dealers <- read_csv("data/atf_dealers.csv")

alb_sf %>%
  ggplot() +
  geom_sf() +
  geom_point(atf_dealers, 
             mapping = aes(x = longitude, y = latitude, col = address_type, shape = address_type), 
             size = 3) +
  scale_color_manual(name = "Address Type", values = c("#F8BE3D", "#007BAB")) +
  scale_shape_manual(name = "Address Type", values = c(17, 19)) +
  theme_void() +
  theme(legend.box.background = element_rect(), legend.box.margin = margin(5, 5, 5, 5)) +
  labs(title = "Firearms Dealers in and Around Albemarle County",
       caption = "Total licensed dealers: 31")
```

## Gun Theft

Police and news outlets have reports on increases in firearm theft, especially from vehicles. 





## References & Resources

Charlottesville and Albemarle:

-   [Cville Community Safety Report](https://prescouncil.president.virginia.edu/sites/g/files/jsddwu616/files/2024-01/Final%20Community%20Safety%20Working%20Group%20Report_24.pdf)

-   [Charlottesville City Health Dashboard](https://www.cityhealthdashboard.com/VA/Charlottesville/city-overview)

National:

-   [KFF: The Impact of Gun Violence on Children and Adolescents](https://www.kff.org/mental-health/issue-brief/the-impact-of-gun-violence-on-children-and-adolescents)

-   [PEW Research: What the Data Says about Gun Deaths in the US](https://www.pewresearch.org/short-reads/2023/04/26/what-the-data-says-about-gun-deaths-in-the-u-s/)

Infrastructure:

-   [Structural Competency and the Future of Firearm Research](https://www.sciencedirect.com/science/article/pii/S0277953621002112)

-   [Racial Equity Impact Assessment Tool for Gun Violence Prevention](https://efsgv.org/wp-content/uploads/WEST_EFSGV_REIA_012275.pdf)

- [AISP: Centering Racial Equity Throughout Data Integration ](https://aisp.upenn.edu/wp-content/uploads/2022/07/AISP-Toolkit_5.27.20.pdf)

- [Science Direct: Effects of "Trauma Recidivism"](https://www.sciencedirect.com/science/article/pii/S0020138322008063#:~:text=Trauma%20recidivism%20was%20defined%20as,treatment%20of%20any%20new%20injury.)
