---
title: "Data Descriptions"
author: Samantha Toet
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(ggmap)
library(janitor)
library(lubridate)
library(magrittr)
library(reactable)
library(tidyverse)

```

The data listed below represent information that is publicly and readily available online. This document is meant to serve as an overview of the existing resources so that we can begin representing the nature, shape, and dimensions of the problem locally.

All of this data is administrative and made up of information created when people interact with public services. As a result, we need to remember the deeply human aspect of this data. Granted personal info is scrubbed from this report, we should keep in mind that these are people and members of our community that we are talking about.

---

### Gun Violence Archive {.tabset}

The national [Gun Violence Archive](https://www.gunviolencearchive.org/) describes gun violence as:

> "all incidents of death or injury or threat with firearms... Violence is defined without intent or consequence as a consideration. To that end a shooting of a victim by a subject/suspect is considered gun violence as is a defensive use or an officer involved shooting. The act itself, no matter the reason is violent in nature."

This database includes information collected from public news sources about 201 incidents of gun violence in the Charlottesville/Albemarle region from May 2014 - Feb 2024. The incidents range from shots fired reports, to unlawful gun ownership, to acts of violence such as assault, suicide, and homicide.

Because the data is sourced from public news websites, if an act of violence didn't receive local news attention it was not included in the Gun Violence Archive database. That means this data underrepresents some information, especially shots fired numbers. Notably, it looks like shots fired incidents haven't been added to the database since 2021.

The data is made up of two dataframes: one has information about each incident and the associated outcome variables (for example, the number of people injured and/or killed), and the other lists the participant-level information (for example, the age rage and gender of the participants for both victims and suspects). The data includes the location and date of each incident along with characteristics such as injury, homicide, or shots fired. The data does not give any identifiable information about the participants, including name, race, ethnicity, socioeconomic status, or mental health status.

#### Incident-level data

```{r gva_incidents}

gva_incidents <- read_csv("data/gva_incidents.csv") 

gva_incidents %>%
  group_by(yr = year(incident_date)) %>%
  summarise(total_injured = sum(victims_injured + suspects_injured),
            total_killed = sum(victims_killed + suspects_killed)) %>%
  ungroup() %>%
  pivot_longer(matches("total")) %>% 
  ggplot(aes(x = yr, y = value, fill = name)) +
  geom_col() +
  scale_fill_discrete(name = "Victim Status",
                      breaks = c("total_injured", "total_killed"),
                      labels = c("Injured", "Killed")) +
  labs(title = "Victims of Gun Violence in the Charlottesville/Albemarle Region",
       x = "Year",
       y = "Number of Victims") +
  theme(axis.text.x = element_text(angle=35))
```

#### Individual-level data

```{r gva_participants}

gva_participants <- read_csv("data/gva_participants.csv") 

gva_participants %>%
  drop_na(age) %>%
  mutate(yr = year(incident_date)) %>%
  ggplot(aes(age, fill = role)) +
  geom_histogram(position = 'identity') +
  facet_grid(~role) +
  scale_fill_discrete(name = "Role",
                      breaks = c("suspect", "victim"),
                      labels = c("Suspect", "Victim")) +
  labs(title = "Ages of Participants",
       x = "Age",
       y = "Number of People") +
  theme(axis.text.x = element_text(angle=35))

```


---

### Charlottesville Open Data Portal {.tabset}

The [Charlottesville Open Data Portal](https://opendata.charlottesville.org/datasets/charlottesville::crime-data/about) includes over 500 gun-related crime reports, and over 400 gun-related arrests in City of Charlottesville from March 2019 through April 2024. The data is managed by the Charlottesville Police Department and updated nightly. 

The dataset is made up of two dataframes: one includes information about initial crime reports and the other includes arrest data. Both datasets have been anonymized to remove names and other identifiable information. 

This is not a definitive collection of public safety data for Charlottesville. We should consider the factors that influence the inclination of an individual calling the police. Are there systematic reasons that someone might call the police more or less than someone else? It's also significant to note that just because a call was made, does not mean that an arrest was made. All individuals are innocent until proven guilty in a court of law.

<i>These datasets only contain information reported by City of Charlottesville Police. It does not contain information about incidents that solely involve other police departments operating within the city (for example, University of Virginia police or Albemarle County police).</i>

```{r odp_crimemap, fig.show="hold", out.width="50%"}

odp_data <- read_csv("data/odp_data.csv") 
arrests <- read_csv("data/arrests.csv")

cville_map <- get_map(c(left = -78.53, bottom = 38.00, right = -78.45, top = 38.07), maptype = "roadmap", color = "bw")

#Reports
ggmap(cville_map) +
  stat_density2d(data = odp_data, aes(fill = ..level.., alpha = 0.1), # removes 55 values
                 geom = "polygon") +
  theme(legend.position="none") +
  scale_fill_viridis_c(direction = -1) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  ggtitle("Gun-Related Crime Reports")

ggmap(cville_map) +
  stat_density2d(data = arrests, aes(fill = ..level.., alpha = 0.1), # removes 96 values
                 geom = "polygon") +
  theme(legend.position="none") +
  scale_fill_viridis_c(direction = -1) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  ggtitle("Gun-Related Arrests")

```

#### Crime Reports 

As defined by the Charlottesville Police Department, a crime report 

>"represents the initial information that is provided by individuals calling for police assistance." 

The crime dataset includes the time of a call, the type of incident, the geocoded location of the incident, and the reporting officer. The data is maintained by CPD and updated nightly. 

The data has been filtered to only include crime reports that involve firearms. To do this, only offences labeled "Shots Fired/Illegal Hunting," "Robbery - Armed," and "Weapons Violations" are used. As a result, this data may significantly underrepresent the number of firearm incidents. For example, crime reports that have "Assault" as the primary offense are not included because we cannot be certain the assault included firearm. 


---

#### Arrest Reports

From the Charlottesville Open Data Portal:

>"Arrest data contains information on people taken into custody by City of Charlottesville police officers. More serious crimes such as felony offenses are more likely to result in an arrest. However, arrests can occur as a result of other offenses, such as parole violations or a failure to appear for trial."

The below data contains the distribution of arrest counts for firearm-related charges. The data has been anonymized so that identifiable information has been removed. Here we can see that of the 233 individuals arrested, 147 only have one gun-related charge, 53 have two charges, and one person has 15 charges. 

```{r odp_arrests}

arrests %>%
  group_by(id) %>%
  summarise(charges = n()) %>%
  ggplot() +
  geom_bar(aes(y = charges)) +
  geom_text(aes(y = charges, label =..count..), stat = "count", hjust = -0.25) +
  labs(title = "Distribution of Gun-Related Arrests",
       x = "Number of People",
       y = "Number of Charges") +
  scale_y_continuous(breaks = c(1:15))
  
```




------------------------------------------------------------------------

### Virginia Open Data Portal {.tabset}

The [Virginia Open Data Portal](https://data.virginia.gov/) includes the number and rate of emergency department (ED) visits for firearm injury (FAI) in Virginia from 2015 - 2024, aggregated by county, health district, and patient demographics. This data is supplied by the Virginia Department of Health and updated monthly. 

Firearm injury visits are identified using key terms in the chief complaint (reason for visit) and discharge diagnosis codes. Key terms include: gun with wound, GSW, gunshot, buckshot, revolver, rifle, shotgun, firearm, pistol, handgun, been shot, I was shot, I got shot, combination of hit, ricochet, graze with bullet. 
ICD-10 diagnosis codes: W32.0, W32.1, W33.0, W33.1, W34.0, W34.1, X72, X73, X74, X93, X94, X95, Y22, Y23, Y24, Y35.0, Y38.4 SNOMED diagnosis codes: 41430008, 56768003, 63409001, 69861004, 77301004, 86122002, 111050005, 219257002, 283545005, 218081007, 218086002, 218082000, 218087006, 218088001, 269796009, 242869008, 219199009, 219200007, 219201006, 219204003, 219205002, 219203009, 219198001, 219142001, 219143006, 219144000, 219145004, 219146003, 287184008, 287193009. 

Exclusions: follow-up visits, visits involving other types of guns (e.g., staple gun), visits where firearm was used as a weapon but not fired (e.g., pistol whip). City/county localities are assigned using the patient's residential zip code for Virginia residents. Patients with non-Virginia or unknown zip codes are grouped as 'Out of State.' 

#### Cville & Albemarle County

```{r vdh_county}

fai_county <- read_csv("data/fai_county.csv") 

fai_county %>%
  filter(str_detect(patient_city_county, "Albemarle")) %>%
  arrange(year) %>%
  select(-patient_city_county) %>%
  reactable()

```

#### Blue Ridge Health District

The Blue Ridge Health District includes Charlottesville City, Albemarle, Greene, Louisa, Fluvanna, and Nelson County. 

```{r vdh_district}

fai_age <- read_csv("data/fai_age.csv")

fai_age %>%
  arrange(age_group) %>%
  reactable(defaultPageSize = 11)
```


